---
title: "lecture-1"
author: "Ammar"
date: "2023-08-27"
output: 
  bookdown::html_document2:
      toc: true
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(bookdown)
```

# Using statnet Package

```{r message=FALSE, warning=FALSE}
library(statnet)
```

## Create a network object from an Adjacency Matrix

```{r}
netmat1 <- rbind(c(0,1,1,0,0),
                 c(0,0,1,1,0),
                 c(0,1,0,0,0),
                 c(0,0,0,0,0),
                 c(0,0,1,0,0))

rownames(netmat1) <- c("A","B","C","D","E")
colnames(netmat1) <- c("A","B","C","D","E")
net1 <- network(netmat1,matrix.type="adjacency")
class(net1)
```

## Summary of 1st Network

```{r}
summary(net1)
```

## Create a network object from an Edge List

```{r}
netmat2 <- rbind(c(1,2),
                 c(1,3),
                 c(2,3),
                 c(2,4),
                 c(3,2),
                 c(5,3))
net2 <- network(netmat2,matrix.type="edgelist")
network.vertex.names(net2) <- c("A","B","C","D","E")
```

## Summary of 2nd Network

```{r}
summary(net2)
```

## Plot Network

```{r}
gplot(net1, vertex.col = 2, displaylabels = TRUE)
```

## Setting vertex (nodes) attributes using base function

```{r}
set.vertex.attribute(net1, "gender", c("F", "F", "M",
                                       "F", "M"))
```

## Using the %v% operator to set a new vertex attribute

```{r}
net1 %v% "alldeg" <- degree(net1)
```

## Getting attributes of network

```{r}
vertex_data <- data.frame(
  vertex.names = get.vertex.attribute(net1, "vertex.names"),
  gender = get.vertex.attribute(net1, "gender"),
  alldeg = get.vertex.attribute(net1, "alldeg")
)

vertex_data

```
## ignore.eval and names.eval

```{r}
netval1 <- rbind(c(0,2,3,0,0),
                 c(0,0,3,1,0),
                 c(0,1,0,0,0),
                 c(0,0,0,0,0),
                 c(0,0,2,0,0))

netval1 <- network(netval1,matrix.type="adjacency",
                   ignore.eval=FALSE,names.eval="like")

get.edge.attribute(netval1, "like")
```

## as.sociomatrix without specifying edge attribute

```{r}
as.sociomatrix(netval1)
```

## as.sociomatrix with specifying edge attribute

```{r}
as.sociomatrix(netval1,"like")
```


- The key here are the ignore.eval and names.eval options. 
- These two options tell the network function to evaluate the actual values in the sociomatrix, and store those values in a new edge attribute called 'like.' 
- Once values are stored in an edge attribute, the original valued matrix can be restored using `as.sociomatrix` coercion function.

## Setting and getting vertex names

```{r}
network.vertex.names(netval1) <- c("A","B","C","D","E")
network.vertex.names(netval1)
```

## detach statnet

```{r message=FALSE, warning=FALSE}
detach(package:statnet)
```

# Using igraph Packages

```{r message=FALSE, warning=FALSE}
library(igraph)
```

## Create a network object from an Adjacency Matrix

```{r}
netmat1 <- rbind(c(0,1,1,0,0),
                 c(0,0,1,1,0),
                 c(0,1,0,0,0),
                 c(0,0,0,0,0),
                 c(0,0,1,0,0))
rownames(netmat1) <- c("A","B","C","D","E")
colnames(netmat1) <- c("A","B","C","D","E")

inet1 <- graph.adjacency(netmat1)
summary(inet1)
```

## structure function

```{r}
str(inet1)
```


## Plot a network with non-weighted edges

```{r}
netmat2 <- rbind(c(1,2),
                 c(1,3),
                 c(2,3),
                 c(2,4),
                 c(3,2),
                 c(5,3))
inet2 <- graph.edgelist(netmat2)

V(inet2)$name <- c("A","B","C","D","E") #vertex names
E(inet2)$val <- c(1:6) #edge width

plot.igraph(inet2,vertex.label=V(inet2)$name,
            edge.width=E(inet2)$weight,edge.color="black",
            layout=layout.fruchterman.reingold)
```

## Plot a network with weighted edges

```{r}
netval2<- rbind(c("A","B",2),
                 c("A","C",1),
                 c("B","C",3),
                 c("B","D",3),
                 c("C","B",2),
                 c("E","C",1))
colnames(netval2) <- c("from","to","like") #setting relationships
inet3 <- graph.edgelist(netval2[,1:2]) #setting vertex
E(inet3)$weight <- netval2[,3] #setting weights

#str(inet3)
plot.igraph(inet3,vertex.label=V(inet3)$name,
            edge.width=E(inet3)$weight,edge.color="black",
            layout=layout.fruchterman.reingold)
```

- Note how for the network with weighted edges, the lines are thicker when the edge's weight > 1.

## detach igraph

```{r}
detach(package:igraph)
```

# Filtering a Network

## using statnet
```{r message=FALSE, warning=FALSE}
library(statnet)
```

```{r}
netmat1 <- rbind(c(0,1,1,0,0),
                 c(0,0,1,1,0),
                 c(0,1,0,0,0),
                 c(0,0,0,0,0),
                 c(0,0,1,0,0))
rownames(netmat1) <- c("A","B","C","D","E")
colnames(netmat1) <- c("A","B","C","D","E")
net1 <- network(netmat1,matrix.type="adjacency")
set.vertex.attribute(net1, "gender", c("F", "F", "M",
                                       "F", "M"))
n1F <- get.inducedSubgraph(net1,
                           which(net1 %v% "gender" == "F")) #filter for females
gplot(n1F,displaylabels=TRUE)
```

- compare this to the network plot earlier when no filtering was conducted. there would be a total of 5 nodes instead of 3

## detach statnet

```{r}
detach(package:statnet)
```

## Using igraph Packages

```{r Load Packages, message=FALSE, warning=FALSE}
library(igraph)
```

```{r}
netval2<- as.data.frame(rbind(c("A","B",2),
                              c("A","C",1),
                              c("B","C",3),
                              c("B","D",3),
                              c("C","B",2),
                              c("E","C",1)))
from=c("A","A","B","B","C","E")
to=c("B","C","C","D","B","C")
like=c(2,1,3,3,2,1)
netval2=data.frame(from,to,like,stringsAsFactors = FALSE)

inet3 <- graph.edgelist(as.matrix(netval2[,1:2]))
E(inet3)$weight <- netval2$like

n2F <- subgraph.edges(inet3,
                      which(E(inet3)$weight>1),
                      delete.vertices = TRUE)  
plot.igraph(n2F,
            vertex.label=V(n2F)$name,
            edge.width=E(n2F)$weight,
            edge.color="black",
            layout=layout.fruchterman.reingold)
```





